---
output: github_document
---

# VALIDICLUST
  <!-- badges: start -->
  [![R-CMD-check](https://github.com/benhvt/VALIDICLUST/workflows/R-CMD-check/badge.svg)](https://github.com/benhvt/VALIDICLUST/actions)
  <!-- badges: end -->

# Overview 

`VALIDICLUST` s a package for ensuring valid inference after data clustering. This problem occurs when clustering forces differences between groups of observations to build clusters. This leads to an inflation of the Type I error rate, especially because the data is used twice: i) to build clusters, i.e., to form hypotheses, and ii) to perform the inference step. The 3 main functions of the package are:

* `test_selective_inference()` following the work of Gao et al. [[1]](#1) on selective inference for clustering

* `merge_selective_inference()` for the merging selective test. In this method, all adjacent p-values are merged using the harmonic mean

* `test_multimod()` uses the DipTest [[2]](#2) to investigate the presence of a continnum between two estimated clusters for a given variable

# Installation

Install the development version from GitHub

```{r, eval = F}
remotes::install_github("benhvt/VALIDICLUST")
```

# Example 

To illustrate our proposed method, we use the [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/) dataset from the `palmerpenguins` package. To ensure that there are no truly separate groups of observations, we selected only the female penguins of the Adelie species, yielding 73 observations. On this dataset, we apply hierarchical clustering (on Euclidean distances with Ward linkage) to build 2 clusters. In this negative control dataset, we know the true, i.e., the non-existing separated group of observations. We therefore apply our 3 tests to each of the 4 numerical measurements to test for separation of the 2 clusters, and compare the resulting p-values to the p-values of the classical t-test.

```{r, echo = F, message = F}
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(GGally)
library(VALIDICLUST)
```

```{r,message = F}
data("penguins")
data <- subset(penguins, species == "Adelie" & sex == "female")
data <- data[,3:6]

# Clustering function 
hcl2 <- function(x){
  x <- scale(x, center = T, scale = T)
  distance <- dist(x, method = "euclidean")
  cah <- hclust(distance, method = "ward.D2")
  return(as.factor(cutree(cah, k=2)))
}

data$Cluster <- hcl2(data)

ggpairs(data, aes(colour = Cluster, fill = Cluster)) + 
  scale_colour_manual(values = c("#F1786D","#53888f")) + 
  scale_fill_manual(values = c("#F1786D","#53888f")) + 
  theme_minimal()

pval.inference.selective <- pval.merge <- pval.multimod <- pval.t.test <- rep(NA, 4)
for (i in 1:4){
  pval.inference.selective[i] <- test_selective_inference(as.matrix(data[,1:4]), 
                                                          k1=1, 
                                                          k2=2, 
                                                          g=i, 
                                                          cl_fun = hcl2, 
                                                          cl=data$Cluster)$pval
  pval.merge[i] <- merge_selective_inference(as.matrix(data[,1:4]),
                                             k1=1, 
                                             k2=2, 
                                             g=i, 
                                             cl_fun = hcl2,
                                             cl=data$Cluster)$pval
  pval.multimod[i] <- test_multimod(as.matrix(data[,1:4]),
                                    g=i,
                                    k1=1, 
                                    k2=2,
                                    cl = data$Cluster)$pval
  pval.t.test[i] <- t.test(data[data$Cluster == 1,i], data[data$Cluster==2, i])$p.value
}


pval.res <- rbind(pval.inference.selective, 
                  pval.merge,
                  pval.multimod,
                  pval.t.test)

colnames(pval.res) <- gsub("_", x=colnames(data)[1:4], replacement = " ")
rownames(pval.res) <- c("Selective Inference",
                        "Merging Selective Inference",
                        "Multimoality test", 
                        "T-test")

round(pval.res,3) %>% htmlTable::htmlTable()
```


# References
<a id="1">[1]</a> 
Gao, L. L., Bien, J., & Witten, D. (2020). Selective inference for hierarchical clustering. arXiv preprint arXiv:2012.02936.

<a id="2">[2]</a>
HARTIGAN, John A. et HARTIGAN, Pamela M. The dip test of unimodality. The annals of Statistics, 1985, p. 70-84.
